---
title: "Aligning SAFMC Policy Objectives and Existing Indicators"
author: "Sarah Gaichas"
date: "`r Sys.Date()`"
output:
  html_document:
    code_fold: hide
    toc: true
    toc_float: true
  pdf_document:
    toc: true
  word_document:
    toc: true
csl: plos.csl
bibliography: operationalSOE.bib
link-citations: true
urlcolor: blue
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
```

# Introduction, Methods, and Caveats

Essential Fish Habitat (EFH) policy documents and current Ecosystem Status Report (ESR) indicators [@craig_ecosystem_2021] were reviewed and aligned where possible, with gaps highlighted. Three key policy documents were reviewed in detail and aligned with existing indicators:

* South Atlantic [Food Webs and Connectivity Policy](https://safmc.net/documents/policy-considerations-for-south-atlantic-food-webs-and-connectivity-and-essential-fish-habitats/) – January 2025
    +  Connects EFH and food web considerations, identifies research needs 
    +  Includes fish food habits data identifying top 10 prey by FMP
    +  Includes food web model for the South Atlantic region
* [Climate Variability and Fisheries](https://safmc.net/wp-content/uploads/document-blocks/2026/01/6bc7f4cd-7c4e-45d3-9c29-fddd42ea8f98-policy-considerations-for-south-atlantic-climate-variability-and-fisheries-and-essential-fish-habitats_march2018.pdf) – March 2018
    +  Connects changing ocean conditions with fish distribution shifts and changing timing, identifies research needs 
    +  Notes these shifts present challenges for management and new fishery opportunities  
    +  Requests annual summaries of climate indicators tracking ecological, social and economic trends 
    +  Requests climate vulnerability analysis and management strategy evaluation focused on climate robustness
* [Marine Submerged Aquatic Vegetation](https://safmc.net/wp-content/uploads/document-blocks/2025/02/6bddb63d-9deb-4fd4-af8c-407d8cd56f17-policy-for-protection-and-enhancement-of-estuarine-and-marine-submerged-aquatic-vegetation-sav-habitat.pdf) - June 2014    
    +  Describes ecosystem services provided by SAV and recent declines in this EFH
    +  Outlines monitoring and research, planning, management, and education and enforcement to protect and restore SAV
    
Additional policies identify species with EFH and HAPC potentially impacted by each activity, and identify specific threats, best management practices and research needs. These may be reviewed for alignment with existing indicators if time permits:  

*  Beach Dredging and Filling, Beach Renourishment and Large-Scale Coastal Engineering – September 2023 
*  Energy Exploration, Development, Transportation and Hydropower Re-Licensing - November 2024   
*  Alteration of Riverine, Estuarine and Nearshore Flows Policy – June 2014    
*  Artificial Reef Habitat Policy Threats to EFH – September 2017    
*  Marine and Estuarine Non-Native and Invasive Species Policy – June 2014    
*  Interactions Between Essential Fish Habitats And Marine Aquaculture Policy – June 2014     

Policy objectives without current indicators were identified. Information gaps that might be filled with Citizen Science programs were then identified, along with what those programs would need to deliver. 

All policy documents were reviewed by the author. Initial matches with ESR indicators were taken from the ESR summary which used Claude Sonnet 4.5 to produce a draft summary and R code to compile indicator descriptions as described [here](https://sgaichas.github.io/SAFMCindicators/ESRcomparisons.html). After initial matches were made, pertinent ESR sections were reviewed by the author to validate matches and provide indicator details. 

# Policy Document Summary

## South Atlantic Food Webs and Connectivity Policy

This document outlines 9 policies and 9 research and information needs. 

Research needs are broader and so considered separate from policies here. Both are aligned with existing indicators in the following tables.

```{r FWpolicy}

FWcrosswalk <- data.frame(Policy = c("Forage Fisheries",
                                     "Prey Importance",
                                     "Food Web Indicators",
                                     "Food Web Connectivity",
                                     "Trophic Pathways",
                                     "Food Web Models",
                                     "Ecosystem Component Species",
                                     "Invasive Species",
                                     "Contaminants"),
                          Brief = c("Consider forage impacts on predator productivity when setting catch limits",
                                    "Classify important prey as ecosystem component species",
                                    "Consider food web targets and thresholds for management action",
                                    "Account for migratory species interactions across otherwise separate food webs",
                                    "Maintain diverse (bottom up and top down) energy pathways",
                                    "Use food web models in multiple decision contexts where appropriate",
                                    "Definition: otherwise unmanaged species important to achieving ecosystem management objectives",
                                    "Account for invasive species impacts in management actions",
                                    "Consider human health and food web impacts"
                                    ),
                          ProposedInd = c("Forage stock abundance and dynamics for important forage by FMP (Appendix A)",
                                    "Mass, occurrence, and degree of overlap among multiple predators",
                                    "Not listed",
                                    "Not listed",
                                    "Diversity of energy pathways",
                                    "Not listed",
                                    "Not listed",
                                    "Lionfish predation on and competition with other reef species",
                                    "Not listed"
                                    ),
                          CurrentInd = c("Forage fish (menhaden) abundance and estuarine crab and shrimp landings",
                                    "Average percent prey in diet by FMP (Appendix A)",
                                    "Mean Trophic Level",
                                    "None",
                                    "None",
                                    "None",
                                    "None",
                                    "None",
                                    "None"
                                    ),
                          PotentialInd = c("Add food web estimated trends for aggregate forage",
                                    "Diet analysis to estimate prey mass, occurrence, and degree of overlap among multiple predators",
                                    "Estimate from food web model",
                                    "Potentially from spatial food web model",
                                    "Estimate from food web model",
                                    "Identify desirable food web states and estimate from food web model",
                                    "Use indicators for Prey Importance",
                                    "Analysis of lionfish diet, distribution and abundance, potentially add to food web model",
                                    "Potentially from seafood safety monitoring"
                                    ),
                          Feasibility = c("High: use current food web model",
                                          "Unknown, dependent on quality of existing food habits data",
                                          "High: use current food web model",
                                          "Moderate: requires spatial update to food web model",
                                          "High: use current food web model",
                                          "High: use current food web model, pending Council ecosystem objectives",
                                          "High: aggregate indicators available",
                                          "Moderate: published lionfish diet studies exist, requires abundance data and model update",
                                          "Low: published contaminant studies have limited spatial scope")

                          )


flextable::flextable(FWcrosswalk) |>
  flextable::width(width = c(1,2,2,1,2,2)) |>
  flextable::set_header_labels(Brief = "Brief Policy Description",
                               ProposedInd = "Indicators Outlined in the Policy",
                               CurrentInd = "Available Indicators",
                               PotentialInd = "Potential Indicators",
                               Feasibility = "Potential Indicator Feasibility") |>
  flextable::set_caption("SAFMC Food Web and Connectivity Policy Objectives and Current Indicators")

```

Overall, while two food web policy related indicators are reported in the 2021 ESR, at least 3 additional policies can be assessed with indicators derived from the existing food web model. Two others could potentially be addressed with more labor intensive updates to the existing food web model by integrating spatial capabilities and existing literature. Some literature information on invasive species diet is available [@peake_feeding_2018; @sancho_invasive_2018]. Contaminant studies are not included in the ESR, perhaps due to limited work in the South Atlantic region [@fair_persistent_2018; @barbieri_contaminants_2025; @thornton_multistressor_2026]. 

```{r FWresearch}

FWResearchCrosswalk <- data.frame(Research = c("Climate Impacts on Productivity",
                                               "Offshore Habitats for Estuarine Species",
                                               "Role of Forage Species",
                                               "Fix Data Gaps",
                                               "Overarching Risks",
                                               "Species Risk Assessments",
                                               "MSEs",
                                               "Ecosystem Reference Points",
                                               "Essential Fish Habitat"),
                                  CurrentInd = c("CVA, Recruitment of Economically Important Species, Coral Bleaching",
                                                 "Surface and Bottom Temperature, FL Current Transport, Gulf Stream Position, Upwelling, Primary Productivity, Ocean Acidification",
                                                 "Forage fish (menhaden) abundance and estuarine crab and shrimp landings",
                                                 "None",
                                                 "CVA, Community CVA, Sea Level Rise, FL Current Transport, Gulf Stream Position, Upwelling",
                                                 "CVA",
                                                 "Peterson et al 2025",
                                                 "None",
                                                 "Wetlands and Forests, SAV, Oyster Reefs, Coral Demographics and Bleaching, Nearshore, Offshore Hard Bottom, and Coral Reef Fish Diversity and Abundance"),
                                  PotentialAction = c("Prioritize and tailor to species based on CVA",
                                                   "Prioritize and tailor existing indicators to species based on CVA",
                                                   "Habitat specific managed species diet",
                                                   "Ensure full use of ESR and CVA data, prioritize data for Citizen Science collection",
                                                   "Consider FMP level risks to refine and align existing indicators with management",
                                                   "Clarify additional factors to build on CVA",
                                                   "List/provide model code, inputs, and outputs across analyses",
                                                   "Develop objectives before reference points",
                                                   "Prioritize and tailor existing indicators to EFH and HAPC"),
                                  Feasibility = c("Moderate: resource intensive depending on priority species",
                                                  "Moderate: resource intensive depending on priority species",
                                                  "Unknown, dependent on quality of existing food habits data",
                                                  "Mixed: will vary by application",
                                                  "High: existing CVA and ESR applied by FMP",
                                                  "Moderate: depends on additional factors needed",
                                                  "High: central repository for existing models",
                                                  "Moderate: requires Council objectives to test literature ELRPs",
                                                  "High: many ESR indicators align with habitats")
  
)

flextable::flextable(FWResearchCrosswalk) |>
  flextable::width(width = c(2,2,2,2)) |>
  flextable::set_header_labels(CurrentInd = "Available Indicators",
                               PotentialAction = "Potential Council Actions",
                               Feasibility = "Potential Action Feasibility") |>
  flextable::set_caption("SAFMC Food Web and Connectivity Research Needs and Current Indicators")


```

Research and information needs are already partially addressed by existing CVA [@burton_climate_2023; @craig_climate_2025] and ESR indicators, and by current MSE efforts [@peterson_climate-readiness_2025]. Collaboration between the Council and NOAA/other partners to identify clear priorities and objectives for research is a practical next step. Many current ESR indicators have the potential to address specific Council needs; collaboration to tailor the indicators for specific uses is recommended. A structured decision process to identify and analyze ecosystem issues (similar to Pacific FEP Initiatives, North Pacific Action Modules, or Gulf Fishery Ecosystem Issues) could help achieve these goals. 

Reference material:

*  [Published lionfish studies](https://coastalscience.noaa.gov/news/lionfish-prefer-to-eat-local/)

## Climate Variability and Fisheries

This document describes historical and projected ocean conditions in the South Atlantic, climate effects on fish, habitats, and fisheries, information gaps and research priorities, and relationships with SAFMC management. Here, stated management policies and research priorities are aligned with existing indicators. This policy was last updated in 2018, before the 2021 ESR and 2023 CVA were released.

```{r ClimPolicy}

Climcrosswalk <- data.frame(Policy = c("Collaboration",
                                     "Climate Indicators",
                                     "Tradeoffs",
                                     "Precautionary",
                                     "New Fisheries"),
                          Brief = c("Work across jurisdictions with multiple organizations and stakeholders",
                                    "Develop and present climate indicators annually",
                                    "Consider increased uncertainty and changing productivity in management",
                                    "Apply precautionary principle under uncertain future climate conditions",
                                    "Manage new fishery development to avoid negative EFH impacts"
                                    ),
                          ProposedInd = c("Species distribution shifts",
                                    "Climate, ecological, social, and economic trends and status, environmentally driven fishery trends",
                                    "Uncertainty and stock productivity",
                                    "Not listed",
                                    "Not listed"
                                    ),
                          CurrentInd = c("None",
                                    "5 Climate, 13 Physical, 5 Habitat, 4 Low and 6 Upper Trophic Level, 8 Fishery, Social, and Economic indicator categories",
                                    "Biomass and Recruitment of Commercially Important Species, Shrimp, Crab, and Oyster landings",
                                    "None",
                                    "None"
                                    ),
                          PotentialInd = c("Latitudinal and depth shifts for all or individual species",
                                    "Review and refine existing before considering additional indicators",
                                    "Short term ocean forecasts including uncertainty, MSE with changing productivity",
                                    "Short term ocean forecasts including uncertainty",
                                    "Landings of currently unmanaged species"
                                    ),
                          Feasibility = c("High: available on DisMAP website",
                                          "High: prioritize available indicators for refinement",
                                          "Moderate: resources required to integrate regional ocean forecasts and uncertainty in MSE",
                                          "Moderate: resources required to process regional ocean forecasts",
                                          "Moderate: resources required to compile information from multiple sources"
                                          )

                          )


flextable::flextable(Climcrosswalk) |>
  flextable::width(width = c(1,2,2,1,2,2)) |>
  flextable::set_header_labels(Brief = "Brief Policy Description",
                               ProposedInd = "Indicators Outlined in the Policy",
                               CurrentInd = "Available Indicators",
                               PotentialInd = "Potential Indicators",
                               Feasibility = "Potential Indicator Feasibility") |>
  flextable::set_caption("SAFMC Climate Variability and Fisheries Policy Objectives and Current Indicators")



```

Many current ESR indicators directly address Climate Variability and Fisheries policy statements. Current gaps related to species distribution shifts and forecast conditions could be filled with existing information. [DisMAP indicators](https://apps-st.fisheries.noaa.gov/dismap/DisMAP.html#) summarizing species distribution from the SEAMAP survey are available. [Regional ocean model forecasts](https://psl.noaa.gov/cefi_portal/) are available for the South Atlantic region, but resources are required to process, evaluate, and translate these outputs into indicators for management review. Some Councils already receive reports of [unmanaged species landings](https://safmc.net/documents/mafmcunmanagedcomlandings-pdf/) compiled by NOAA Regional Offices. However, these reports focus on commercial landings because there are few information sources for recreational landings of unmanaged species.

```{r ClimResearch}

ClimateResearchCrosswalk <- data.frame(Research = c("Climate Impacts on Productivity",
                                               "Climate Impacts in Assessments",
                                               "3D Estuarine-Coastal-Ocean Observations",
                                               "Climate Robust MSE",
                                               "Socioeconomic Impacts and Fishery Responses",
                                               "Offshore Habitats for Estuarine Species"),
                                  CurrentInd = c("CVA combined with many ESR indicators",
                                                 "ESR indicators need refinement for assessment application",
                                                 "Surface and Bottom Temperature, FL Current Transport, Gulf Stream Position, Upwelling",
                                                 "Peterson et al 2025",
                                                 "CVA, Community CVA, Sea Level Rise, FL Current Transport, Gulf Stream Position, Upwelling",
                                                 "Surface and Bottom Temperature, FL Current Transport, Gulf Stream Position, Upwelling, Primary Productivity, Ocean Acidification"),
                                  PotentialAction = c("Prioritize and tailor existing indicators to species based on CVA",
                                                   "Prioritize and tailor existing ESR indicators to assessments based on CVA",
                                                   "Add ocean model reanalysis products (GLORYS) to existing indicators",
                                                   "Consider standard climate scenarios in all MSEs",
                                                   "Prioritize and tailor existing indicators",
                                                   "Prioritize and tailor existing indicators to species based on CVA"),
                                  Feasibility = c("Moderate: resource intensive depending on priority species",
                                                  "Moderate: resource intensive depending on priority species",
                                                  "High: apply methods and code developed in other regions",
                                                  "High: develop simple climate scenarios as robustness tests",
                                                  "High: existing CVA, Community CVA, and ESR indicators",
                                                  "Moderate: resource intensive depending on priority species")
  
)

flextable::flextable(ClimateResearchCrosswalk) |>
  flextable::width(width = c(2,2,2,2)) |>
  flextable::set_header_labels(CurrentInd = "Available Indicators",
                               PotentialAction = "Potential Council Actions",
                               Feasibility = "Potential Action Feasibility") |>
  flextable::set_caption("SAFMC Climate Variability and Fisheries Research Needs and Current Indicators")



```

There are multiple ways to move forward with Climate Variability and Fisheries policies and research priorities given existing indicators and analyses. An initial step is collaborative Council and NOAA review of current indicators to prioritize a subset for refinement and annual updates. The CVA results could help identify a subset of species or a representative species for each FMP for initial focus. Considerable effort towards understanding the social and economic context and community climate vulnerability already exists [@jepson_development_2013; @seara_community_2022] and can contribute to these efforts. Note that the last research priority here is identical to the second research priority identified in the Food Webs and Connectivity Policy.

## Marine Submerged Aquatic Vegetation

Submerged Aquatic Vegetation (SAV) or seagrass is an important habitat in the South Atlantic, found primarily in North Carolina and Florida coastal waters. SAFMC has designated SAV to be Essential Fish Habitat (EFH) for managed fish (snapper-grouper species and cobia) and invertebrates (Penaeid shrimp and spiny lobster), as well as a Habitat Area of Particular Concern (HAPC) for snapper-grouper species. 

The Council does not directly manage many of the activities and conditions that affect SAV. The overall policy goal in this document is to support actions that protect and restore SAV throughout the South Atlantic. These actions require collaboration with the four South Atlantic states: North Carolina, South Carolina, Georgia, and Florida, as well as local authorities. 

```{r SAVPolicy}

SAVcrosswalk <- data.frame(Policy = c("Monitoring",
                                     "Planning",
                                     "Management",
                                     "Education and Enforcement"),
                          Brief = c("Monitoring is needed for assessment and management",
                                    "Establish goals, objectives, measures of success",
                                    "Review existing human activity rules, water quality standards, and restoration guidelines",
                                    "Analyze and communicate benefits of SAV protection, evaluate current enforcement"
                                    ),
                          ProposedInd = c("SAV distribution and shifts, water quality",
                                    "Not listed",
                                    "Not listed",
                                    "Not listed"
                                    ),
                          CurrentInd = c("SAV extent and % change",
                                    "SAV extent and % change, Coastal salinity",
                                    "Coastal salinity, Stream flow, Nutrient loading, Coastal and Urban land use",
                                    "Total ocean economy"
                                    ),
                          PotentialInd = c("Same indicator with more comprehensive data in space and time",
                                    "Change in SAV distribution; habitat depth, sediment, light penetration, salinity, and wave energy",
                                    "Habitat disturbance indicators (dredging, construction, bottom contact from boating and fishing)",
                                    "Commercial and Recreational landings and value for SAV dependent species"
                                    ),
                          Feasibility = c("Low-Moderate: depends on multiple states resources and coordination",
                                          "Mixed: depends on water quality information availability",
                                          "Low-Moderate: resource intensive to collate information from many local sources",
                                          "High: subset of information from current indicators"
                                          )

                          )


flextable::flextable(SAVcrosswalk) |>
  flextable::width(width = c(1,2,2,1,2,2)) |>
  flextable::set_header_labels(Brief = "Brief Policy Description",
                               ProposedInd = "Indicators Outlined in the Policy",
                               CurrentInd = "Available Indicators",
                               PotentialInd = "Potential Indicators",
                               Feasibility = "Potential Indicator Feasibility") |>
  flextable::set_caption("SAFMC Submerged Aquatic Vegetation Policy Objectives and Current Indicators")


```

Existing ESR indicators for SAV, coastal habitat attributes, land use and ocean economy align with SAV policy.  Additional indicators of SAV benefits to dependent managed species could be derived from current indicators for commercial and recreational landings and value. However, comprehensive SAV indicators and research would require dedicated effort from multiple partners. 

```{r SAVResearch}

SAVResearchCrosswalk <- data.frame(Research = c("Standardize Mapping Protocols",
                                               "EFH GIS Database",
                                               "Evaluate Water Quality",
                                               "Drivers of SAV Loss",
                                               "Restoration Efficacy",
                                               "Climate Impacts on SAV"),
                                  CurrentInd = c("None",
                                                 "SAV areal coverage and % change based on existing data",
                                                 "Stream Flow, Nutrient Loading, Precipitation and Drought",
                                                 "SAV, Stream Flow, Nutrient Loading, Precipitation and Drought, Sea Level Rise, Storms and Hurricanes, Primary Productivity, Human Population, Coastal Land Use",
                                                 "None",
                                                 "SAV, Surface and Bottom Temperature, Primary Productivity, Ocean Acidification"),
                                  PotentialAction = c("Collaborate with partners to design and implement regional surveys",
                                                   "Expand Council EFH and HAPC maps",
                                                   "Prioritize existing indicators for further refinement",
                                                   "Prioritize existing indicators for further refinement",
                                                   "Collaborate with partners",
                                                   "Consider habitat climate vulnerability assessment"),
                                  Feasibility = c("Moderate: resource intensive proces",
                                                  "Moderate: resource intensive process",
                                                  "Moderate-High: may need additional indicators for water clarity",
                                                  "High: many potential indicators for SAV drivers exist",
                                                  "Moderate: resource intensive process",
                                                  "Moderate: resource intensive process")
  
)

flextable::flextable(SAVResearchCrosswalk) |>
  flextable::width(width = c(2,2,2,2)) |>
  flextable::set_header_labels(CurrentInd = "Available Indicators",
                               PotentialAction = "Potential Council Actions",
                               Feasibility = "Potential Action Feasibility") |>
  flextable::set_caption("SAFMC Submerged Aquatic Vegetation Research Needs and Current Indicators")



```

In general, research priorities for SAV require more coordination and collaboration with state and local partners relative to the priorities above for food web connectivity and climate fisheries interactions. However, the research objectives are clearly stated and well focused. Initial steps could include evaluation and prioritization of ESR indicators for relevance to SAV extent and % change. A longer term project could consider habitat-climate vulnerability analysis as was done for Northeast US Atlantic habitats [@farr_assessment_2021], which would apply not just to SAV but to all Council designated EFH and HAPC.

May contain some useful information but more marsh focused: https://www.conservationgateway.org/collections/land/resilient-coastal-sites-for-conservation-in-the-south-atlantic-us/

# References

