---
title: "SAFMC Policy Objectives and Existing Indicators"
author: "Sarah Gaichas"
date: "`r Sys.Date()`"
output:
  html_document:
    code_fold: hide
    toc: true
    toc_float: true
  pdf_document:
    toc: true
  word_document:
    toc: true
csl: plos.csl
bibliography: operationalSOE.bib
link-citations: true
urlcolor: blue
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
```

# Introduction, Methods, and Caveats

EFH policy documents and current ESR indicators [@craig_ecosystem_2021] were reviewed and aligned where possible, with gaps highlighted. Three key policy documents were reviewed in detail and aligned with existing indicators:

* South Atlantic [Food Webs and Connectivity Policy](https://safmc.net/documents/policy-considerations-for-south-atlantic-food-webs-and-connectivity-and-essential-fish-habitats/) – January 2025
    +  Connects EFH and food web considerations, identifies research needs 
    +  Includes fish food habits data identifying top 10 prey by FMP
    +  Includes food web model for the South Atlantic region (Fig. \@ref(fig:SAfoodweb))  
* Climate Variability and Fisheries – December 2016
    +  Connects changing ocean conditions with fish distribution shifts and changing timing, identifies research needs 
    +  Notes these shifts present challenges for management and new fishery opportunities  
    +  Requests annual summaries of climate indicators tracking ecological, social and economic trends 
    +  Requests climate vulnerability analysis and management strategy evaluation focused on climate robustness
* Marine Submerged Aquatic Vegetation (SAV) Habitat Policy – June 2014    
    +  Describes ecosystem services provided by SAV and recent declines in this EFH
    +  Outlines monitoring and research, planning, management, and education and enforcement to protect and restore SAV
    
Additional policies identify species with EFH and HAPC potentially impacted by each activity, and identify specific threats, best management practices and research needs. These were reviewed for alignment with existing indicators as time permitted:  

*  Beach Dredging and Filling, Beach Renourishment and Large-Scale Coastal Engineering – September 2023 
*  Energy Exploration, Development, Transportation and Hydropower Re-Licensing - November 2024   
*  Alteration of Riverine, Estuarine and Nearshore Flows Policy – June 2014    
*  Artificial Reef Habitat Policy Threats to EFH – September 2017    
*  Marine and Estuarine Non-Native and Invasive Species Policy – June 2014    
*  Interactions Between Essential Fish Habitats And Marine Aquaculture Policy – June 2014     

Policy objectives without current indicators were identified. Information gaps that might be filled with Citizen Science programs were then identified, along with what those programs would need to deliver. 

All policy documents were reviewed by the author. Initial matches with ESR indicators were taken from the ESR summary which used Claude Sonnet 4.5 to produce a draft summary and R code to compile indicator descriptions as described here. After initial matches were made, pertinent ESR sections were reviewed by the author to validate matches and provide indicator details. 

# Policy Document Summary

## South Atlantic Food Webs and Connectivity Policy

This document outlines 9 policies and 9 research and information needs. 

Research needs are broader and so considered separate from policies here. Both are aligned with existing indicators in the following tables.

```{r FWpolicy}

FWcrosswalk <- data.frame(Policy = c("Forage Fisheries",
                                     "Prey Importance",
                                     "Food Web Indicators",
                                     "Food Web Connectivity",
                                     "Trophic Pathways",
                                     "Food Web Models",
                                     "Ecosystem Component Species",
                                     "Invasive Species",
                                     "Contaminants"),
                          Brief = c("Consider forage impacts on predator productivity when setting catch limits",
                                    "Classify important prey as ecosystem component species",
                                    "Consider food web targets and thresholds for management action",
                                    "Account for migratory species interactions across otherwise separate food webs",
                                    "Maintain diverse (bottom up and top down) energy pathways",
                                    "Use food web models in multiple decision contexts where appropriate",
                                    "Definition: otherwise unmanaged species important to achieving ecosystem management objectives",
                                    "Account for invasive species impacts in management actions",
                                    "Consider human health and food web impacts"
                                    ),
                          ProposedInd = c("Forage stock abundance and dynamics for important forage by FMP (Appendix A)",
                                    "Mass, occurrence, and degree of overlap among multiple predators",
                                    "Not listed",
                                    "Not listed",
                                    "Diversity of energy pathways",
                                    "Not listed",
                                    "Not listed",
                                    "Lionfish predation on and competition with other reef species",
                                    "Not listed"
                                    ),
                          CurrentInd = c("Forage fish (menhaden) abundance and estuarine crab and shrimp landings",
                                    "Average percent prey in diet by FMP (Appendix A)",
                                    "Mean Trophic Level",
                                    "None",
                                    "None",
                                    "None",
                                    "None",
                                    "None",
                                    "None"
                                    ),
                          PotentialInd = c("Add food web estimated trends for aggregate forage",
                                    "Diet analysis to estimate prey mass, occurrence, and degree of overlap among multiple predators",
                                    "Estimate from food web model",
                                    "Potentially from spatial food web model",
                                    "Estimate from food web model",
                                    "Identify desirable food web states and estimate from food web model",
                                    "Use indicators for Prey Importance",
                                    "Analysis of lionfish diet, distribution and abundance, potentially add to food web model",
                                    "Potentially from seafood safety monitoring"
                                    ),
                          Feasibility = c("High: use current food web model",
                                          "Unknown, dependent on quality of existing food habits data",
                                          "High: use current food web model",
                                          "Moderate: requires spatial update to food web model",
                                          "High: use current food web model",
                                          "High: use current food web model, pending Council ecosystem objectives",
                                          "High: aggregate indicators available",
                                          "Moderate: published lionfish diet studies exist, requires abundance data and model update",
                                          "Low: published contaminant studies have limited spatial scope")

                          )


flextable::flextable(FWcrosswalk) |>
  flextable::width(width = c(1,2,2,1,2,2)) 

```

Overall, while two food web policy related indicators are reported in the 2021 ESR, at least 3 additional policies can be assessed with indicators derived from the existing food web model. Two others could potentially be addressed with more labor intensive updates to the existing food web model by integrating spatial capabilities and existing literature. Some literature information on invasive species diet is available [@peake_feeding_2018; @sancho_invasive_2018]. Contaminant studies are not included in the ESR, perhaps due to limited work in the South Atlantic region [@fair_persistent_2018; @barbieri_contaminants_2025; @thornton_multistressor_2026]. 

```{r FWresearch}

FWResearchCrosswalk <- data.frame(Research = c("Climate Impacts on Productivity",
                                               "Offshore Habitats for Estuarine Species",
                                               "Role of Forage Species",
                                               "Fix Data Gaps",
                                               "Overarching Risks",
                                               "Species Risk Assessments",
                                               "MSEs",
                                               "Ecosystem Reference Points",
                                               "Essential Fish Habitat"),
                                  CurrentInd = c("CVA, Recruitment of Economically Important Species, Coral Bleaching",
                                                 "Surface and Bottom Temperature, FL Current Transport, Gulf Stream Position, Upwelling, Primary Productivity, Ocean Acidification",
                                                 "Forage fish (menhaden) abundance and estuarine crab and shrimp landings",
                                                 "None",
                                                 "CVA, Community CVA, Sea Level Rise, FL Current Transport, Gulf Stream Position, Upwelling",
                                                 "CVA",
                                                 "Peterson et al 2024",
                                                 "None",
                                                 "Wetlands and Forests, SAV, Oyster Reefs, Coral Demographics and Bleaching, Nearshore, Offshore Hard Bottom, and Coral Reef Fish Diversity and Abundance"),
                                  PotentialAction = c("Prioritize and tailor to species based on CVA",
                                                   "Prioritize and tailor existing indicators to species based on CVA",
                                                   "Habitat specific managed species diet",
                                                   "Ensure full use of ESR and CVA data, prioritize data for Citizen Science collection",
                                                   "Consider FMP level risks to refine and align existing indicators with management",
                                                   "Clarify additional factors to build on CVA",
                                                   "List/provide model code, inputs, and outputs across analyses",
                                                   "Develop objectives before reference points",
                                                   "Prioritize and tailor existing indicators to EFH and HAPC"),
                                  Feasibility = c("Moderate: resource intensive depending on priority species",
                                                  "Moderate: resource intensive depending on priority species",
                                                  "Unknown, dependent on quality of existing food habits data",
                                                  "Mixed: will vary by application",
                                                  "High: existing CVA and ESR applied by FMP",
                                                  "Moderate: depends on additional factors needed",
                                                  "High: central repository for existing models",
                                                  "Moderate: requires Council objectives to test literature ELRPs",
                                                  "High: many ESR indicators align with habitats")
  
)

flextable::flextable(FWResearchCrosswalk) |>
  flextable::width(width = c(2,2,2,2)) 


```

Research and information needs are already partially addressed by existing CVA and ESR indicators, and by current MSE efforts [@peterson_climate-readiness_2025]. Collaboration between the Council and NOAA/other partners to identify clear priorities and objectives for research is a practical next step. Many current ESR indicators have the potential to address specific Council needs; collaboration to tailor the indicators for specific uses is recommended. A structured decision process to identify and analyze ecosystem issues (similar to Pacific FEP Initiatives, North Pacific Action Modules, or Gulf Fishery Ecosystem Issues) could help achieve these goals. 

Reference material:

*  [Published lionfish studies](https://coastalscience.noaa.gov/news/lionfish-prefer-to-eat-local/)

## Climate Variability and Fisheries



# References

