---
title: "Prototype Ocean Indicators"
author: "Sarah Gaichas"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_fold: hide
    df_print: paged
    toc: true
    toc_float: true
  bookdown::pdf_document2:
    includes:
      in_header: latex/header1.tex
    keep_tex: true
    toc: false
    number_sections: false
  pdf_document:
    toc: true
  word_document:
    toc: true
  html_document:
    code_fold: hide
csl: plos.csl
bibliography: operationalSOE.bib
link-citations: true
urlcolor: blue
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
reticulate::use_miniconda("/Users/sarahgaichas/Library/r-miniconda-arm64")
```

# Introduction

Ocean data is available for the Southeast US coast, from multiple sources. Regional ocean models and global ocean models, including the NOAA MOM6 model [@ross_dynamically_2024] (https://psl.noaa.gov/cefi_portal/) and the Copernicus products (https://data.marine.copernicus.eu/products) including the  Global Oceanic Reanalysis (GLORYS, [@jean-michel_copernicus_2021], https://data.marine.copernicus.eu/product/GLOBAL_MULTIYEAR_PHY_001_030/description) already provide both hindcasts and forecasts of ocean temperature and salinity for the South Atlantic. This project will identify specific resources necessary to translate available data into ecosystem indicators tailored for the South Atlantic and its fishery management needs. For example, the NEFSC has publicly available code to derive indicators from these data sources for the region north of Cape Hatteras that could be modified to derive prototype indicators for the South Atlantic region. Here, I develop prototype ocean indicators to illustrate both opportunities and challenges for operational application in the region.

A prototype habitat indicator characterizing bottom temperature in space over time was developed using GLORYS data with a reproducible workflow based on workflows used in the Northeast US by NEFSC. The objective is to produce maps and time series indicating where and when bottom temperature conditions approach or exceed stressful levels input particular species ranging from corals to targeted fish species. Stressful temperature levels would need to be determined from literature or other studies if unknown. 

Validation of this model-based temperature indicator could be achieved with an expanded Citizen Science program; establishment of a formal Study Fleet such as that in the Northeast US [@jones_learning_2022] will be outlined if the indicator shows promise.  

# Methods

## Pull GLORYS bottom temperature

To use GLORYS information, users must register for a free account on the Copernicus website by clicking the [link at the top of the page](https://data.marine.copernicus.eu/register). Users supply an email and optional information on their organization (one can select Others) and then receive an email with a link to establish a password.

I am working in R [@r_core_team_r_2023], but [Python](https://docs.python.org/3/index.html) code is used to access GLORYS information (there is not a way to access in R). Python can be installed using the R package [`reticulate`](https://rstudio.github.io/reticulate/) [@ushey_reticulate_2025] and the command `reticulate::install_miniconda()` if you don't already have a Python installation. This installs [Miniconda](https://www.anaconda.com/docs/getting-started/miniconda/main), a lightweight distribution of Python. Because it is a minimal distribution, one or two packages need to be added as below. 

For this script you need the python `pandas` package, which is installed using `reticulate::py_install("pandas")`

*These lines are left here for the record but I was unable to use the`motuclient` package*
You also need the `motuclient` package, from https://github.com/clstoulouse/motu-client-python 
Try `reticulate::py_install("motuclient", pip = TRUE)` which worked. Almost. `reticulate::py_install("setuptools", pip = TRUE)`
Maybe motuclient doesnt work anymore?
*I also didn't need to install with pip, which was recommended for `motuclient` but not needed for other packages, I kept using it to stay consistent*
*The lines below would work, possibly better, without the pip="TRUE" argument*

We also need to install the package from Copernicus that makes downloading the information easy:

Install the `copernicusmarine` package using `reticulate::py_install("copernicusmarine", pip = TRUE)`

Install `h5py`. `reticulate::py_install("h5py", pip = TRUE)`

The dataset id we want is `cmems_mod_glo_phy_my_0.083deg_P1D-m`

With the username and password, the following code can be used to download data locally for processing. It is important to specify the latitude and longitude bounds of the data as well as the year(s). Downloading one year at a time makes for manageable files to process for spatial or time series indicators. The output directory is hardcoded below, this needs to change for use elsewhere.

This code for downloading is courtesy Joseph Caracappa, NOAA NMFS NEFSC, and is a test downloading NEUS information so I can compare my results with published indicators to ensure my code is working correctly:

```{python testNE, eval=FALSE}

import datetime
import pandas
import os
from os import path

#Set start and end year
year_start=2020
year_end=2024
years=range(year_start,year_end+1) 

#Input CMEMS User and Password (case sensitive)
USER = ""
PASSWORD = ""

#Set Lat/lon bounds
min_lon = str(-82.5)
max_lon = str(-51.5)
min_lat = str(22.5)
max_lat = str(48.5)

for y in range(len(years)):

    out_dir = "...ADD YOUR PATH HERE.../SAFMCindicators/GLORYS/testNE/"+str(years[y])+"/"
    if(not os.path.exists(out_dir)):
        os.makedirs(out_dir)
        
        
    dt = datetime.datetime(years[y],1,1)
    end = datetime.datetime(years[y],12,31)
    step = datetime.timedelta(days = 1)
    
    all_days = []
    
    while dt <= end:
        all_days.append(dt.strftime('%Y-%m-%d'))
        dt += step
        
    for d in range(len(all_days)):
        t1 = all_days[d]+"T00:00:00"
        t2 = all_days[d]+"T23:59:59"
        #Change to desired filename prefix
        new_name = "GLORYS_REANALYSIS_"+all_days[d]+".nc"
        #Change to appropriate output path
        
        
        if(path.exists(out_dir+new_name)):
           print(new_name+" EXISTS")
           continue
           
        #print(new_name)
        #If additional variables are desired: need to add "--variable var.name"

        command = "copernicusmarine subset -i cmems_mod_glo_phy_my_0.083deg_P1D-m -x "+min_lon+" -X "+max_lon+" -y "+min_lat+" -Y "+max_lat+" -z 0. -Z 5000. -t "+t1+" -T "+t2+" -v sea_water_potential_temperature_at_sea_floor -o "+out_dir+" -f "+new_name+" --force-download --username "+USER+" --password "+PASSWORD
        #command = "copernicusmarine subset -i cmems_mod_glo_phy_myint_0.083deg_P1D-m -x "+min_lon+" -X "+max_lon+" -y "+min_lat+" -Y "+max_lat+" -z 0. -Z 5000. -t "+t1+" -T "+t2+" -v thetao -o "+out_dir+" -f "+new_name+" --force-download"
        
        #print(command)
        os.system(command)

```

There are separate files for each day of each year.

A single file called `GLORYS_REANALYSIS_2025-12-31.nc` is now downloaded to the local GLORYS/testNE/2025 folder. 

Take a look at it

```{r}
tidync::tidync(here::here("GLORYS/testNE/2025/GLORYS_REANALYSIS_2025-12-31.nc"))
```

To highlight a box for the South Atlantic region, I visited https://data.marine.copernicus.eu/product/GLOBAL_MULTIYEAR_PHY_001_030/download then selected variables, drew a box around area on the map, selected a year range, then clicked "automate" to generate a command line script. This script was modified and put into the same framework as used above for the Northeast region. 

This script pulled initial bottom temperature files for the South Atlantic, 2020-2025:

```{python SAtl, eval=FALSE}

import datetime
import pandas
import os
from os import path

#Set start and end year
year_start=2020
year_end=2025
years=range(year_start,year_end+1) 

#Input CMEMS User and Password (case sensitive)
USER = ""
PASSWORD = ""

#Set Lat/lon bounds
min_lon = str(-81.9)
max_lon = str(-73.6)
min_lat = str(23.9)
max_lat = str(36.6)

for y in range(len(years)):

    out_dir = "...ADD YOUR PATH HERE.../GLORYS/SAtl/"+str(years[y])+"/"
    if(not os.path.exists(out_dir)):
        os.makedirs(out_dir)
        
        
    dt = datetime.datetime(years[y],1,1)
    end = datetime.datetime(years[y],12,31)
    step = datetime.timedelta(days = 1)
    
    all_days = []
    
    while dt <= end:
        all_days.append(dt.strftime('%Y-%m-%d'))
        dt += step
        
    for d in range(len(all_days)):
        t1 = all_days[d]+"T00:00:00"
        t2 = all_days[d]+"T23:59:59"
        #Change to desired filename prefix
        new_name = "GLORYS_REANALYSIS_"+all_days[d]+".nc"
        #Change to appropriate output path
        
        
        if(path.exists(out_dir+new_name)):
           print(new_name+" EXISTS")
           continue
           
        #print(new_name)
        #If additional variables are desired: need to add "--variable var.name"

        command = "copernicusmarine subset -i cmems_mod_glo_phy_my_0.083deg_P1D-m -x "+min_lon+" -X "+max_lon+" -y "+min_lat+" -Y "+max_lat+" -z 0. -Z 5000. -t "+t1+" -T "+t2+" -v sea_water_potential_temperature_at_sea_floor -o "+out_dir+" -f "+new_name+" --force-download --username "+USER+" --password "+PASSWORD
        #command = "copernicusmarine subset -i cmems_mod_glo_phy_myint_0.083deg_P1D-m -x "+min_lon+" -X "+max_lon+" -y "+min_lat+" -Y "+max_lat+" -z 0. -Z 5000. -t "+t1+" -T "+t2+" -v thetao -o "+out_dir+" -f "+new_name+" --force-download"
        
        #print(command)
        os.system(command)

```

Pulling 6 years of data took my laptop and home connection about 8 hours. 

Take a look at a South Atlantic file:

```{r}
tidync::tidync(here::here("GLORYS/SAtl/2025/GLORYS_REANALYSIS_2025-12-31.nc"))
```

## Process GLORYS bottom temperature

Joe Caracappa, NEFSC, wrote nearly all of the code used to produce these indicators for the Northeast US [State of the Ecosystem](https://www.fisheries.noaa.gov/new-england-mid-atlantic/ecosystems/state-ecosystem-reports-northeast-us-shelf) reports. 

R packages used in this workflow include `tidyverse` [@wickham_welcome_2019] for general data organization and visualization, `terra` [@hijmans_terra_2025] for processing netCDF files downloaded from Copernicus, and `knitr` [@xie_knitr_2025; @xie_dynamic_2015 ; @stodden_knitr_2014], `rmarkdown` [@allaire_rmarkdown_2024; @xie_r_2018; @xie_r_2020], and `bookdown` [@xie_bookdown_2025; @xie_bookdown_2016] for producing reports in both .html and .pdf formats.

Code for reading the file into time series or maps is found at the NEFSC GitHub

https://github.com/NEFSC/READ-EDAB-GLORYS/tree/main

This repo has a [fuller version of the pulling script](https://github.com/NEFSC/READ-EDAB-GLORYS/blob/main/data-raw/python/GLORYS_download_template_workflow.py) that adds variables like salinity or mixed layer depth and pulls them into variable specific folders. It could be modified simply by changing the min and max lat and lon as above.

It also has functions to process the downloaded files into datasets for presentation as either time series or spatial datasets. It calls utility functions in this repository:

https://github.com/NEFSC/READ_EDAB_Utilities/tree/main

Finally, functions to plot indicators are found in the `ecodata` [@beltz_ecodata_2025] package and in the NOAA-EDAB GitHub ecodata repository, such as `plot_thermal_habitat_gridded` : https://github.com/NOAA-EDAB/ecodata/blob/pre-production/R/plot_thermal_habitat_gridded.R 

First we try producing the NEUS indicators to validate the workflow:

```{r}

```

Next we try the same code on the South Atlantic GLORYS bottom temperature.



# References

