---
title: "Prototype Ocean Indicators"
author: "Sarah Gaichas"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_fold: hide
    df_print: paged
    toc: true
    toc_float: true
  bookdown::pdf_document2:
    includes:
      in_header: latex/header1.tex
    keep_tex: true
    toc: false
    number_sections: false
  pdf_document:
    toc: true
  word_document:
    toc: true
csl: plos.csl
bibliography: operationalSOE.bib
link-citations: true
urlcolor: blue
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
reticulate::use_miniconda("/Users/sarahgaichas/Library/r-miniconda-arm64")
```

# Introduction

Ocean data is available for the Southeast US coast, from multiple sources. Regional ocean models and global ocean models, including the NOAA MOM6 model [@ross_dynamically_2024] (https://psl.noaa.gov/cefi_portal/) and the Copernicus products (https://data.marine.copernicus.eu/products) including the  Global Oceanic Reanalysis (GLORYS, [@jean-michel_copernicus_2021], https://data.marine.copernicus.eu/product/GLOBAL_MULTIYEAR_PHY_001_030/description) already provide both hindcasts and forecasts of ocean temperature and salinity for the South Atlantic. This project will identify specific resources necessary to translate available data into ecosystem indicators tailored for the South Atlantic and its fishery management needs. For example, the NEFSC has publicly available code to derive indicators from these data sources for the region north of Cape Hatteras that could be modified to derive prototype indicators for the South Atlantic region. Here, I develop prototype ocean indicators to illustrate both opportunities and challenges for operational application in the region.

A prototype habitat indicator characterizing bottom temperature in space over time was developed using GLORYS data with a reproducible workflow based on workflows used in the Northeast US by NEFSC. The objective is to produce maps and time series indicating where and when bottom temperature conditions approach or exceed stressful levels input particular species ranging from corals to targeted fish species. Stressful temperature levels would need to be determined from literature or other studies if unknown. 

Validation of this model-based temperature indicator could be achieved with an expanded Citizen Science program; establishment of a formal Study Fleet such as that in the Northeast US [@jones_learning_2022] will be outlined if the indicator shows promise.  

# Methods

To use GLORYS information, users must register for a free account on the Copernicus website by clicking the [link at the top of the page](https://data.marine.copernicus.eu/register). Users supply an email and optional information on their organization (one can select Others) and then receive an email with a link to establish a password.

I am working in R and Rstudio, but Python code is used to access GLORYS information (there is not a way to access in R). Python can be installed using the R package [`reticulate`](https://rstudio.github.io/reticulate/) and the command `reticulate::install_miniconda()` if you don't already have a Python installation. 

For this script you need the python `pandas` package, which is installed using `reticulate::py_install("pandas")`

You also need the motuclient package, from https://github.com/clstoulouse/motu-client-python 

Try `reticulate::py_install("motuclient", pip = TRUE)` which worked. Almost. `reticulate::py_install("setuptools", pip = TRUE)`

Maybe motuclient doesnt work anymore?

Try `copernicusmarine`.  `reticulate::py_install("copernicusmarine", pip = TRUE)`

Install `h5py`. `reticulate::py_install("h5py", pip = TRUE)`

Dataset id `cmems_mod_glo_phy_my_0.083deg_P1D-m`

With the username and password, the following code can be used to download data locally for processing. It is important to specify the latitude and longitude bounds of the data as well as the year(s). Downloading one year at a time makes for manageable files to process for spatial or time series indicators. The output directory is hardcoded below, this needs to change for use elsewhere.

This code for downloading is courtesy Joseph Caracappa, NOAA NMFS NEFSC:

```{python testNE, eval=FALSE}

import datetime
import pandas
import os
from os import path

#Set start and end year
year_start=2020
year_end=2024
years=range(year_start,year_end+1) 

#Input CMEMS User and Password (case sensitive)
USER = "sgaichas"
PASSWORD = "ZAQ!2wsxCDE#4rfv"

#Set Lat/lon bounds
min_lon = str(-82.5)
max_lon = str(-51.5)
min_lat = str(22.5)
max_lat = str(48.5)

for y in range(len(years)):

    out_dir = "/Users/sarahgaichas/Documents/Work/SAFMCindicators/SAFMCindicators/GLORYS/testNE/"+str(years[y])+"/"
    if(not os.path.exists(out_dir)):
        os.makedirs(out_dir)
        
        
    dt = datetime.datetime(years[y],1,1)
    end = datetime.datetime(years[y],12,31)
    step = datetime.timedelta(days = 1)
    
    all_days = []
    
    while dt <= end:
        all_days.append(dt.strftime('%Y-%m-%d'))
        dt += step
        
    for d in range(len(all_days)):
        t1 = all_days[d]+"T00:00:00"
        t2 = all_days[d]+"T23:59:59"
        #Change to desired filename prefix
        new_name = "GLORYS_REANALYSIS_"+all_days[d]+".nc"
        #Change to appropriate output path
        
        
        if(path.exists(out_dir+new_name)):
           print(new_name+" EXISTS")
           continue
           
        #print(new_name)
        #If additional variables are desired: need to add "--variable var.name"

        command = "copernicusmarine subset -i cmems_mod_glo_phy_my_0.083deg_P1D-m -x "+min_lon+" -X "+max_lon+" -y "+min_lat+" -Y "+max_lat+" -z 0. -Z 5000. -t "+t1+" -T "+t2+" -v sea_water_potential_temperature_at_sea_floor -o "+out_dir+" -f "+new_name+" --force-download --username "+USER+" --password "+PASSWORD
        #command = "copernicusmarine subset -i cmems_mod_glo_phy_myint_0.083deg_P1D-m -x "+min_lon+" -X "+max_lon+" -y "+min_lat+" -Y "+max_lat+" -z 0. -Z 5000. -t "+t1+" -T "+t2+" -v thetao -o "+out_dir+" -f "+new_name+" --force-download"
        
        #print(command)
        os.system(command)

```

There are separate files for each day of each year.

A single file called `GLORYS_REANALYSIS_2025-12-31.nc` is now downloaded to the local GLORYS/testNE/2025 folder. 

Take a look at it

```{r}
tidync::tidync(here::here("GLORYS/testNE/2025/GLORYS_REANALYSIS_2025-12-31.nc"))
```

This script should pull bottom temperature for the South Atlantic:

```{python SAtl, eval=FALSE}

import datetime
import pandas
import os
from os import path

#Set start and end year
year_start=2020
year_end=2025
years=range(year_start,year_end+1) 

#Input CMEMS User and Password (case sensitive)
USER = "sgaichas"
PASSWORD = "ZAQ!2wsxCDE#4rfv"

#Set Lat/lon bounds
min_lon = str(-81.9)
max_lon = str(-73.6)
min_lat = str(23.9)
max_lat = str(36.6)

for y in range(len(years)):

    out_dir = "/Users/sarahgaichas/Documents/Work/SAFMCindicators/SAFMCindicators/GLORYS/SAtl/"+str(years[y])+"/"
    if(not os.path.exists(out_dir)):
        os.makedirs(out_dir)
        
        
    dt = datetime.datetime(years[y],1,1)
    end = datetime.datetime(years[y],12,31)
    step = datetime.timedelta(days = 1)
    
    all_days = []
    
    while dt <= end:
        all_days.append(dt.strftime('%Y-%m-%d'))
        dt += step
        
    for d in range(len(all_days)):
        t1 = all_days[d]+"T00:00:00"
        t2 = all_days[d]+"T23:59:59"
        #Change to desired filename prefix
        new_name = "GLORYS_REANALYSIS_"+all_days[d]+".nc"
        #Change to appropriate output path
        
        
        if(path.exists(out_dir+new_name)):
           print(new_name+" EXISTS")
           continue
           
        #print(new_name)
        #If additional variables are desired: need to add "--variable var.name"

        command = "copernicusmarine subset -i cmems_mod_glo_phy_my_0.083deg_P1D-m -x "+min_lon+" -X "+max_lon+" -y "+min_lat+" -Y "+max_lat+" -z 0. -Z 5000. -t "+t1+" -T "+t2+" -v sea_water_potential_temperature_at_sea_floor -o "+out_dir+" -f "+new_name+" --force-download --username "+USER+" --password "+PASSWORD
        #command = "copernicusmarine subset -i cmems_mod_glo_phy_myint_0.083deg_P1D-m -x "+min_lon+" -X "+max_lon+" -y "+min_lat+" -Y "+max_lat+" -z 0. -Z 5000. -t "+t1+" -T "+t2+" -v thetao -o "+out_dir+" -f "+new_name+" --force-download"
        
        #print(command)
        os.system(command)

```


Code for reading the file into time series or maps is found at the NEFSC github

https://github.com/NEFSC/READ-EDAB-GLORYS/tree/main

https://github.com/NEFSC/READ_EDAB_Utilities/tree/main



```{r}

```


# References

