---
title: "GLORYS reanalysis bottom temperature"
author: "Sarah Gaichas"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_fold: hide
    df_print: paged
    toc: true
    toc_float: true
  bookdown::pdf_document2:
    includes:
      in_header: latex/header1.tex
    keep_tex: true
    toc: false
    number_sections: false
  pdf_document:
    toc: true
  word_document:
    toc: true
  html_document:
    code_fold: hide
csl: plos.csl
bibliography: operationalSOE.bib
link-citations: true
urlcolor: blue
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(patchwork)

```

## GLORYS bottom temperature at the scale of the SADL survey

Copernicus products (https://data.marine.copernicus.eu/products) including the  Global Oceanic Reanalysis (GLORYS, [@jean-michel_copernicus_2021], https://data.marine.copernicus.eu/product/GLOBAL_MULTIYEAR_PHY_001_030/description) provide both hindcasts and forecasts of ocean temperature and salinity for the full Atlantic spanning the Northeast and Southeast regions.

Bottom temperature is reported as the daily mean in this dataset. 

Information was pulled from the Copernicus website using code in https://sgaichas.github.io/SAFMCindicators/PrototypeOceanIndicator.html#21_Pull_GLORYS_bottom_temperature 

There are separate files for each day of each year.

Lets look at July 20 2025, one year ago. 

A single file called `GLORYS_REANALYSIS_2025-07-20.nc` is now downloaded to the local GLORYS/testNE/2025 folder. 

Take a look at it

```{r}
tidync::tidync(here::here("GLORYS/testNE/2025/GLORYS_REANALYSIS_2025-07-20.nc"))
```

Plot it at the full extent pulled

```{r}

ncfile <- here::here("GLORYS/testNE/2025/GLORYS_REANALYSIS_2025-07-20.nc")

# find the unit from https://ropensci.org/blog/2019/11/05/tidync/
tunit <- ncmeta::nc_atts(ncfile, "time") %>% tidyr::unnest(cols = c(value)) %>% dplyr::filter(name == "units")
# its hours since 1950-01-01

glorysnctotibble <- function(ncfile = ncfile){
  # time is hours since 1950-01-01
  origin <- as.Date("1950-01-01") 
  
  # add the time variable into each
  time <- tidync::tidync(ncfile) |>
    tidync::activate("D0") |> tidync::hyper_array()
  
 # google says 86400000 milliseconds in a day
  # irrelevant because this unit is hours in a day
  
  bttib <- tidync::tidync(ncfile) |>
    tidync::hyper_tibble(force = TRUE) |>
    dplyr::mutate(time = unname(time$time)) |>
    dplyr::mutate(date = as.Date((time/24), origin = origin),
                  year = lubridate::year(date),
                  month = lubridate::month(date),
                  day = lubridate::day(date))|>
    dplyr::mutate(longitude = as.numeric(longitude),
                  latitude = as.numeric(latitude)) |>
    dplyr::rename(mod_bt = bottomT)
  
  return(bttib)
}

dailybtplot <- function(oneday, xlim = c(-82.5, -65), ylim = c(23.5, 45)){
  ggplot() +
    geom_tile(data = oneday, aes(x = longitude, y = latitude, fill = mod_bt)) +
    geom_sf(data = ecodata::coast) +
    #geom_point(data = FishStatsUtils::northwest_atlantic_grid, aes(x = Lon, y = Lat), size=0.05, alpha=0.1) +
    scale_fill_gradientn(name = "Temp C",
                         limits = c(0.5, 31),
                         colours = c(scales::muted("blue"), "white",
                                     scales::muted("red"), "black")) +
    #coord_sf(xlim = c(-77, -65), ylim = c(35, 45)) + 
    #oord_sf(xlim = c(-82.5, -71), ylim = c(23.5, 36)) + 
    coord_sf(xlim = xlim, ylim = ylim) + 
    ecodata::theme_map() +
    ggtitle(paste("Bottom temp, mm-dd-yyyy:", unique(oneday$month),
                   unique(oneday$day), unique(oneday$year), sep = " "))
}

jul2025 <- glorysnctotibble(ncfile)

dailybtplot(jul2025)

```


Zoom in on the SADL area

```{r}



dailybtplot(jul2025,
            xlim = c(-80, -72),
            ylim = c(30, 40))

```


What does September 20 look like?

```{r}

ncfile <- here::here("GLORYS/testNE/2025/GLORYS_REANALYSIS_2025-09-20.nc")

sep2025 <- glorysnctotibble(ncfile)

dailybtplot(sep2025,
            xlim = c(-80, -72),
            ylim = c(30, 40))

```

Really zoom in off Hatteras and VA

```{r}

j <- dailybtplot(jul2025,
            xlim = c(-77, -73),
            ylim = c(34, 39)) 

s <- dailybtplot(sep2025,
            xlim = c(-77, -73),
            ylim = c(34, 39)) 


j + s
```

Really zoom in off Hatteras and VA

```{r}

j <- dailybtplot(jul2025,
            xlim = c(-77, -74),
            ylim = c(35, 37)) 

s <- dailybtplot(sep2025,
            xlim = c(-77, -74),
            ylim = c(35, 37)) 


j + s
```

One degree each dimension, now you can see the 1/12 degree resolution of the bottom temperature information.

```{r}

j <- dailybtplot(jul2025,
            xlim = c(-76, -75),
            ylim = c(35, 36)) 

s <- dailybtplot(sep2025,
            xlim = c(-76, -75),
            ylim = c(35, 36)) 


j + s
```

This information is available daily for the period of the SADL survey 2023-2025.

## References

